\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-12-20 main.tex
push everything x-mas

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

Evolutionary game theory~\cite{smith1982evolution, hofbauer1998evolutionary,
nowak2004evolutionary, hauert2005game} describes the evolutionary dynamics of
populations consisting of different types of interacting individuals. The
framework of evolutionary game theory has been applied in %ToDo cite.
Traditional approaches of evolutionary game theory assume that individuals meet
each other at random in an infinitely large well-mixed population, such an
approach is the replicator dynamics. The replicator dynamics describes how the
abundance of strategic types in a population changes based on their fitness. In
this deterministic formulation, individuals with higher fitness increase in
abundance and ultimately, the system reaches a stable fixed point in which the
population may consist either of a single type or of a mixture of different
types. The works of cite have shown that by constrain the population to a
finiteness of populations may lead to fundamental changes in this picture due to
stochastic effects. These stochastic effects disadvantageous mutants have a
small, yet non-zero probability to reach fixation in a finite population.

Two classes of such finite stochastic processes have been used extensively: (i)
fitness-based processes in which an individual chosen proportional to fitness
reproduces and the offspring replaces a randomly chosen individual [11]
\textit{moran process}; (ii) \textit{pairwise comparison processes} in which a
pair of individuals is chosen, and where subsequently one of these individuals
may adopt the strategy of the other.

In pairwise comparison process considers a finite population of fixed size. In
the population at a time step different types of individuals can exist. In the
simplest of cases each individual can be of one of two types, A and B. The state
of the population is thus characterized by the number \(k\) of individuals of
type A. The interaction between the two types of individuals is described by the
functions \(\pi _{{\rm A}}^{i}\) and $\pi _{{\rm B}}^{i}$.

At each step an individual change their type based.

\begin{equation} \label{Eq:rho}
  \rho(\pi _{{\rm A}}^{i}, \pi _{{\rm B}}^{i}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (\pi _{{\rm A}}^{i}\!-\!\pi _{{\rm B}}^{i})}}.
\end{equation}

The parameter denotes the intensity of selection. For , fitness equals payoff.
This scenario describes ``strong selection''. For \(\beta \ll 1\) , the
payoff only provides a small perturbation to the overall fitness of an
individual, a limit known as weak selection (Nowak et al., 2004).

These indicate the expected payoff for two types in a population in state i. The
interaction can be generated from a two-player matrix game which leads to
payoffs linear in i [26], but we keep the formalism general to include games
played between an arbitrary number of players, which leads to payoffs that
depend on i in a polynomial way [27–30]. In fact, our results hold for an
arbitrary dependence of the payoffs on i.

In the simplest case, the payoffs of A and B individuals only depend on the
fraction of both types in the population. If there are i A individuals and B
individuals, then the A and B individuals have payoffs and , respectively.
Self-interactions are excluded.

Here, we consider a process based on pairwise comparison between individuals.
Two individuals, A and B, are selected at random. The individual chosen for
reproduction A replaces B with probability p, which depends on the payoff
difference between the two individuals. The composition of the population can
only change if both individuals are of different types. We follow (Blume, 1993,
Szabó and Tőke, 1998, Hauert and Szabó, 2005) in choosing the Fermi function
from statistical physics for p

To explain the evolution of cooperation remains one of the greatest problems for
biological and social sciences. Cooperation is the action of choosing to help
others at one's own expense. The standard game of formulating such situation is
the Prisoner's Dilemma~\cite{trivers1971evolution,
milinski1987tit, glynatsi2021bibliometric, ohtsuki2006simple}, in which two
players can choose to cooperate or to defect. The players are offered a certain
payoff, \(R\), for mutual cooperation and a lower payoff, \(P\), for mutual
defection. If one player cooperates while the other defects, then the cooperator
gets the lowest payoff, \(S\), while the defector gains the highest payoff,
\(T\). Thus, the payoffs have the following property \(T > R > P > S\), making
defection the dominant strategy in the non-repeated game.

In the context of the prisoner's dilemma the payoffs are calculated as follows.
In the example that a strategy is playing always cooperate and always defect,

This model has several assumptions. Two assumptions are that initally it is
assumed that time to interact is faster than the time of the population
update. Secondly it is assumed that the indiviail can interact will every
other population member, and recall their interaction perfecttly.

The first assumption was questioned by the work of.

In this work

@@

To investigate the role of extortion in the context of evolutionary games, we
concentrate on the donation game

In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population and we
compare this to the classical scenario of the expected payoffs. We assume that
each pair of players interacts in a donation game. The donation game is a
special case of the prisoner's dilemma. Each player can choose to cooperate by
providing a benefit \(b\) to the other player at their cost \(c\), with \(0 < c
< b\). Thus, the feasible payoffs in each round are \(\mathcal{U} = \{b-c, -c,
b, 0\}\).

@@

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \vspace{.3cm}
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different selection strength values.}
We vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
probabilities \(p, q\) for resident population over \(10^7\) time steps for each
\(\beta\) value. ({\bf B}) The cooperation rate within the resident population
(averaged over all games and over all time steps) over \(\beta\). Unless
explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$,
$c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
$t\!=\!5\times 10^7$ time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

Evolutionary game theory~\cite{smith1982evolution, hofbauer1998evolutionary,
nowak2004evolutionary, hauert2005game} describes the evolutionary dynamics of
populations consisting of different types of interacting individuals. The
framework of evolutionary game theory has been applied in explaining the
behavior of sticklebacks~\cite{milinski1987tit}, the rock-scissors-paper cycles
in bacterial populations~\cite{kerr2002local} and food sharing in vampire
bats~\cite{carter2020development} colonies.

Traditional approaches of evolutionary game theory assume that individuals meet
each other at random in an infinitely large well-mixed population, an example of
such an approach is the replicator dynamics. The replicator dynamics describes
how the abundance of strategic types in a population changes based on their
fitness. In this deterministic formulation, individuals with higher fitness
increase in abundance and ultimately, the system reaches a stable fixed point in
which the population may consist either of a single type or of a mixture of
different types. The works of~\cite{nowak2004emergence, taylor2004evolutionary,
imhof2005evolutionary, imhof2006evolutionary, traulsen2007pairwise} have shown
that constraining the population to be finite can lead to fundamental changes
due to stochastic effects. Due to these stochastic effects disadvantageous
mutants have a small, yet non-zero probability to reach fixation in a finite
population.

Two classes of such finite stochastic processes have been used extensively: (i)
fitness-based processes in which an individual chosen proportional to fitness
reproduces and the offspring replaces a randomly chosen individual~\cite{nowak2004emergence}
\textit{moran process}; (ii) \textit{pairwise comparison processes} in which a
pair of individuals is chosen, and where subsequently one of these individuals
may adopt the strategy of the other~\cite{traulsen2007pairwise}.

A pairwise comparison process considers a finite population of fixed size. In
the population at a time step different types of individuals can exist. In the
simplest of cases each individual can be of one of two types, A and B. The state
of the population is thus characterized by the number \(k\) of individuals of
type A. The interaction between the two types of individuals is described by the
functions \(\pi _{{\rm A}}^{i}\) and $\pi _{{\rm B}}^{i}$.

At each step an individual change their type based. Two individuals, A and B,
are selected at random. The individual chosen for reproduction A replaces B with
probability \(\rho\), which depends on the fitness difference \(\pi _{{\rm
A}}^{i} - \pi _{{\rm B}}^{i}\) between the two individuals. The composition of
the population can only change if both individuals are of different types. The
probability \(\rho\) follows the Fermi function,

\begin{equation}
  \rho(\pi _{{\rm A}}^{i}, \pi _{{\rm B}}^{i}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (\pi _{{\rm A}}^{i}\!-\!\pi _{{\rm B}}^{i})}}.
\end{equation}

The parameter \(\beta\) denotes the intensity of selection.

The fitness of the individuals (\(\pi _{{\rm A}}^{i}, \pi _{{\rm B}}^{i}\)) is
identified with the payoff resulting from the underline game. The evolution of
cooperation remains one of the greatest problems for biological and social
sciences. Cooperation is the action of choosing to help others at one's own
expense. The standard game of formulating such situation is the Prisoner's
Dilemma~\cite{trivers1971evolution, milinski1987tit, glynatsi2021bibliometric,
ohtsuki2006simple}, in which two players can choose to cooperate or to defect.
The players are offered a certain payoff, \(R\), for mutual cooperation and a
lower payoff, \(P\), for mutual defection. If one player cooperates while the
other defects, then the cooperator gets the lowest payoff, \(S\), while the
defector gains the highest payoff, \(T\). Thus, the payoffs have the following
property \(T > R > P > S\), making defection the dominant strategy in the
non-repeated game.

\begin{equation}\label{eq:payoffs}
  \begin{blockarray}{ccc}
      & \text{cooperate} & \text{defect} \\
      \begin{block}{c(cc)}
          \text{cooperate} & R & S \\
          \text{defect} & T & P \\
      \end{block}
  \end{blockarray}
\end{equation}

In literature, the payoffs of A and B individuals only on the fraction of both
types in the population. If there are \(k\) A individuals and \(N-k\) B
individuals. In the context of the prisoner's dilemma the payoffs and assuming
that A and B follow simple strategies where they cooperate and defect
unconditionally respective, then the A and B individuals have fitness of

\begin{equation}
  \begin{array}{lcrcr}
  \displaystyle \pi _{{\rm A}}^{i}	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot R	&+	&\displaystyle\frac{k}{N-1}\cdot S,\\[0.5cm]
  \displaystyle \pi _{{\rm B}}^{i}&=	&\displaystyle\frac{N-k}{N-1}\cdot T &+	&\displaystyle\frac{k-1}{N-1}\cdot P.\\
  \end{array}
\end{equation}

respectively while self-interactions are excluded. In literature these are the
expected payoffs. For \(\beta = 1\), fitness equals payoff. This scenario
describes ``strong selection''. For \(\beta \ll 1\) , the payoff only provides a
small perturbation to the overall fitness of an individual, a limit known as
weak selection.

In the evolutionary game theory literature it is usually assumed that players
use strategies with finite memory~\cite{Nowak1992tit, Baek2016}. This assumption
is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. Players' payoffs are calculated in the limit
of interactions assuming that the two strategies meet several times.

Customarily, most evolutionary game studies make the additional assumption that
individuals play many times and with all other players before reproduction takes
place, so that payoffs, equivalently fitness, are given by the mean distribution
of types in the population. This hypothesis, selection occurs much more slowly
than the interaction between individuals and more that remember all the
interactions they participated in. Although recent experimental studies show
that this may not always be the case in biology [12– 14], it is clear that in
cultural evolution or social learning the time scale of selection is much closer
to the time scale of interaction. The effects of this mixing of scales cannot be
disregarded [15], and then it is natural to ask about the consequences of the
above assumption and the effect of relaxing it. The work of~\cite{Roca2006}.

In this work we ask studies that make use of this classical framework report
that cooperation can substantially evolve. However, the inconsistency in the
memory size of players leads us to question the robustness of our understanding
of cooperation. To this end, we propose a framework in which individuals,
similar to the decisions at each turn, estimate their fitness based on a minimum
of information.

@@

To investigate the role of extortion in the context of evolutionary games, we
concentrate on the donation game.
In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population and we
compare this to the classical scenario of the expected payoffs. We assume that
each pair of players interacts in a donation game. The donation game is a
special case of the prisoner's dilemma. Each player can choose to cooperate by
providing a benefit \(b\) to the other player at their cost \(c\), with \(0 < c
< b\). Thus, the feasible payoffs in each round are \(\mathcal{U} = \{b-c, -c,
b, 0\}\).

@@

\begin{figure}[!htbp]
  \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
\caption{{\bf The evolution of cooperation for different selection strength values.}
We vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
probabilities \(p, q\) for resident population over \(10^7\) time steps for each
\(\beta\) value. ({\bf B}) The cooperation rate within the resident population
(averaged over all games and over all time steps) over \(\beta\). Unless
explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$,
$c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
$t\!=\!5\times 10^7$ time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

\begin{figure}[!htbp]
  \centering
    \includegraphics[width=\textwidth]{static/more_memory.pdf}
% \caption{{\bf The evolution of cooperation for different selection strength values.}
% We vary the selection strength $\beta$. In all cases, stochastic payoff
% evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
% probabilities \(p, q\) for resident population over \(10^7\) time steps for each
% \(\beta\) value. ({\bf B}) The cooperation rate within the resident population
% (averaged over all games and over all time steps) over \(\beta\). Unless
% explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$,
% $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
% $t\!=\!5\times 10^7$ time steps for each parameter
% combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}
